% =============================================================================
% Chapter 3: Behavioural Feature Design
% =============================================================================

\chapter{Behavioural Feature Design}
\label{ch:features}

This section presents the 32 behavioural features used to characterise algorithm
trajectories.  Eleven features are inherited from the BLADE framework
\cite{vanStein2025BLADE, vanStein2025Behaviour}. The remaining 21 are introduced
by this thesis.  Of those 21, six are entirely novel metrics that have not
appeared in prior optimisation literature.  These novel features constitute a
central contribution of this work. They capture aspects of search dynamics (momentum, burstiness, plateau structure, and dimensional heterogeneity) that existing metrics overlook.

Every feature is computed from a single trajectory. A trajectory is represented by a DataFrame of $T = 10{,}000$
function evaluations in $d = 5$ dimensions with bounds $[-5, 5]$.  Columns
consist of the evaluation index, the raw objective value~$y_t$, and the decision
variables $x_t^{(0)}, \ldots, x_t^{(d-1)}$.  We denote the best-so-far fitness
at evaluation~$t$ as $f^*_t = \min_{s \le t} y_s$.

% ============================================================================
\section{Design Principles}
\label{sec:features:principles}

Four criteria guided the selection and design of behavioural features.

\begin{enumerate}
  \item \textbf{Computational cheapness.} Feature extraction must not dominate experiment runtime. A single MA-BBOB evaluation already costs 30--120\,s (running the candidate algorithm 50 times across 10 problems with 5 instances each). Behavioural features should add at most a small fraction of that budget.

  \item \textbf{Distinctness.} Each feature should capture a dimension of behaviour not already covered by other features. We enforce this during feature selection by removing any feature whose Pearson correlation with a higher-ranked feature exceeds $|\rho| > 0.8$.

  \item \textbf{Interpretability for LLM feedback.} Features are reported to the language model as part of the behavioural feedback string. They must therefore carry clear, concise semantics that an LLM can act on. For instance, \emph{``your step sizes are erratic''} is actionable, while \emph{``your Lempel--Ziv complexity is 0.73''} is not, unless accompanied by a description.

  \item \textbf{Discrimination.} Features should separate high-performing from low-performing algorithms. We quantify this via Spearman correlation ~\cite{spearman1904proof} with AOCC and Kolmogorov--Smirnov effect size ~\cite{kolmogorov1933} between the top-25\% and bottom-25\% candidates.
\end{enumerate}

% ============================================================================
\section{Existing BLADE Metrics}
\label{sec:features:blade}

BLADE ships with eleven behavioural metrics introduced by van~Stein et
al.~\cite{vanStein2025Behaviour, vanStein2025BLADE}. We summarise them here and
refer the reader to the original work for formal definitions.  The metrics fall
into four functional categories.

\paragraph{Exploration and diversity.}
\begin{itemize}
  \item \texttt{avg\_nearest\_neighbor\_distance}: mean Euclidean distance between consecutive evaluated points, quantifying how much new territory the search covers at each step.
  \item \texttt{dispersion}: maximum gap in search-space coverage via KD-tree. Large values indicate unvisited regions.
  \item \texttt{avg\_exploration\_pct}: chunk-based measure of
  within-chunk variance relative to a random baseline.
\end{itemize}

\paragraph{Exploitation and intensification.}
\begin{itemize}
  \item \texttt{avg\_distance\_to\_best}: mean distance from each evaluation to the best-so-far solution.
  \item \texttt{intensification\_ratio}: fraction of evaluations within a fixed radius of the final best solution. High values indicate strong local exploitation.
  \item \texttt{avg\_exploitation\_pct}: complement of
  \texttt{avg\_exploration\_pct}, measuring the fraction of
  within-chunk variance attributable to local refinement.
\end{itemize}

\paragraph{Convergence.}
\begin{itemize}
  \item \texttt{average\_convergence\_rate}: geometric mean of successive best-so-far error ratios.
  \item \texttt{avg\_improvement}: mean fitness gain on steps that update the best-so-far.
  \item \texttt{success\_rate}: fraction of all evaluations that improve the best-so-far.
\end{itemize}

\paragraph{Stagnation.}
\begin{itemize}
  \item \texttt{longest\_no\_improvement\_streak}: length of the longest consecutive run without best-so-far improvement. Streaks of several thousand out of $T = 10{,}000$ indicate the algorithm has effectively stopped progressing.
  \item \texttt{last\_improvement\_fraction}: how late in the budget the final improvement occurs, normalised to $[0, 1]$.
\end{itemize}

% ============================================================================
\section{Step-Size and Movement Dynamics}
\label{sec:features:stepsize}

The first group of new features characterises the basic mechanics of how a
candidate algorithm moves through decision space.  All four are standard
trajectory analysis measures catalogued in the meta-feature survey by Cenikj et
al ~\cite{cenikj2026features}. Let $s_t = \lVert x_{t+1} - x_t \rVert$ denote the Euclidean step size at
evaluation~$t$.

\paragraph{\texttt{step\_size\_mean}.}
The mean step size across the trajectory:
\begin{equation}
\label{eq:step_mean}
  \bar{s} = \frac{1}{T-1}\sum_{t=1}^{T-1} s_t.
\end{equation}
Large values of the mean step size~$\bar{s}$ indicate broad exploration.
Small values indicate local exploitation.

\paragraph{\texttt{step\_size\_std}.}
The standard deviation of step sizes:
\begin{equation}
\label{eq:step_std}
  \sigma_s = \sqrt{\frac{1}{T-2}\sum_{t=1}^{T-1}
    \bigl(s_t - \bar{s}\bigr)^2}.
\end{equation}
High step-size standard deviation~$\sigma_s$ signals alternating
exploration and exploitation phases. Low variability signals a uniform
step-size regime throughout the run.

\paragraph{\texttt{step\_size\_trend}.}
The ordinary least-squares slope of step size against evaluation index:
\begin{equation}
\label{eq:step_trend}
  \beta_s
    = \frac{\sum_{t}(t - \bar{t})(s_t - \bar{s})}
           {\sum_{t}(t - \bar{t})^2}.
\end{equation}
A negative step-size trend~$\beta_s$ indicates the classic convergence
pattern of contracting step sizes. A positive slope suggests the
algorithm diverges over time.

\paragraph{\texttt{directional\_persistence}.}
The mean cosine similarity between consecutive displacement vectors
$d_t = x_{t+1} - x_t$:
\begin{equation}
\label{eq:dir_persist}
  \phi = \frac{1}{T-2}\sum_{t=2}^{T-1}
    \frac{d_t \cdot d_{t-1}}
         {\lVert d_t \rVert \, \lVert d_{t-1} \rVert}.
\end{equation}
Values of the directional persistence~$\phi$ near~$+1$ indicate that the
search maintains a consistent heading
(gradient-following behaviour). Values near~$0$ indicate a random walk.
Negative values indicate anti-persistent zigzag motion.

% ============================================================================
\section{Information-Theoretic Features}
\label{sec:features:infotheo}

The four features in this category apply established time-series complexity
measures to the raw fitness sequence $\{y_1, \ldots, y_T\}$.  While each
measure is well-known in signal processing and physics, their application to
optimisation algorithm fitness trajectories on BBOB benchmarks is new.

\paragraph{\texttt{fitness\_sample\_entropy}.}
Sample Entropy (SampEn) quantifies the regularity of the fitness
series~\cite{richman2000sampen}.  Given embedding dimension~$m$ and tolerance
$r = 0.2 \cdot \mathrm{std}(y)$, it counts the conditional probability that
sequences matching for $m$~consecutive values still match at length $m+1$:
\begin{equation}
\label{eq:sampen}
  \mathrm{SampEn}(m, r)
    = -\ln \frac{A^m(r)}{B^m(r)},
\end{equation}
where $B^m$ is the number of template matches of length~$m$ and $A^m$ the
number that also match at length $m+1$ (excluding self-matches).  Low SampEn
indicates a repetitive, self-similar convergence pattern. High SampEn indicates
complex, unpredictable dynamics.  We use $m = 2$ and subsample every 10th
evaluation ($N = 1{,}000$) to reduce the $O(N^2 m)$ cost from ${\sim}30$\,s to
${\sim}0.3$\,s per trajectory.

\paragraph{\texttt{fitness\_permutation\_entropy}.}
Normalised Permutation Entropy (PE) maps the fitness series to a sequence of
ordinal patterns and measures their distributional uniformity~\cite{bandt2002permutation}.
Each window of $D$~consecutive values is replaced by its rank permutation
$\pi \in S_D$, and PE is the Shannon entropy of the permutation distribution
normalised by $\log_2(D!)$:
\begin{equation}
\label{eq:pe}
  \mathrm{PE}(D)
    = -\frac{1}{\log_2(D!)}\sum_{\pi \in S_D} p(\pi) \log_2 p(\pi).
\end{equation}
We use $D = 5$, giving $D! = 120$ possible patterns against $T = 10{,}000$
data points.  $\mathrm{PE} \approx 0$ corresponds to perfectly monotone
convergence. $\mathrm{PE} \approx 1$ to a completely random sequence.

\paragraph{\texttt{fitness\_autocorrelation}.}
The lag-1 autocorrelation of the raw fitness series, following Weinberger's
framework for fitness landscape correlation structure~\cite{weinberger1990fitness}:
\begin{equation}
\label{eq:fitac}
  \rho_y(1)
    = \frac{\sum_{t=1}^{T-1}(y_t - \bar{y})(y_{t+1} - \bar{y})}
           {\sum_{t=1}^{T}(y_t - \bar{y})^2}.
\end{equation}
Importantly, while Weinberger's autocorrelation has long been used to
characterise landscapes via \emph{separate} random walks, it has not previously
been computed on the algorithm's \emph{own} trajectory on BBOB.  High positive
$\rho_y(1)$ indicates smooth, locally exploitative search. Values near zero
indicate the algorithm jumps between unrelated fitness regions.

\paragraph{\texttt{fitness\_lempel\_ziv\_complexity}.}
Normalised Lempel--Ziv complexity (LZC) measures the rate at which new
patterns appear in a binarised version of the fitness
series~\cite{lempel1976complexity}.  The binarisation maps each step to
$b_t = \mathbf{1}[y_{t+1} < y_t]$ (improvement or not), yielding a binary
string.  LZ76 parsing counts the number of distinct substrings~$c(N)$, and the
normalised complexity is:
\begin{equation}
\label{eq:lzc}
  \mathrm{LZC}_n = \frac{c(N)}{N / \log_2 N}.
\end{equation}
$\mathrm{LZC}_n \approx 1$ indicates a random improvement/stagnation pattern.
$\mathrm{LZC}_n \approx 0$ indicates highly repetitive structure (e.g.,
long unbroken stagnation followed by a burst of improvements).

% ============================================================================
\section{Adapted Population Dynamics}
\label{sec:features:dynamo}

Cenikj et al.~\cite{cenikj2023dynamorep} introduced DynamoRep, a set of
trajectory-based features that characterise how a population's spatial and
fitness distributions change across generations.  Their features assume a
population-based algorithm with multiple individuals per generation.  In our
$(1{+}1)$-ES setting the population size is~1, so per-generation statistics
are undefined.  We adapt the DynamoRep idea by comparing the first 25\% of
evaluations (\emph{early}) with the last 25\% (\emph{late}), treating each
window as a proxy for the population at that stage of the search.

Let $X_{\mathrm{early}} = \{x_t : t \le T/4\}$ and
$X_{\mathrm{late}} = \{x_t : t > 3T/4\}$.

\paragraph{\texttt{x\_spread\_early} and \texttt{x\_spread\_late}.}
Mean per-dimension standard deviation of decision variables in the early and
late windows, respectively:
\begin{equation}
\label{eq:spread}
  S_{\mathrm{E}}
    = \frac{1}{d}\sum_{j=0}^{d-1}
      \mathrm{std}\!\bigl(\{x_t^{(j)} : t \le T/4\}\bigr).
\end{equation}
The late spread~$S_{\mathrm{L}}$ is defined identically over $t > 3T/4$.
High early spread~$S_{\mathrm{E}}$ indicates broad initial exploration.
Low~$S_{\mathrm{L}}$ indicates the search has converged to a small
region.

\paragraph{\texttt{spread\_ratio}.}
The ratio of late to early spread:
\begin{equation}
\label{eq:spread_ratio}
  R_S = S_{\mathrm{L}} \;/\; S_{\mathrm{E}}.
\end{equation}
Values of the spread ratio~$R_S$ below~1 indicate the algorithm narrows
its search over time (the expected convergence pattern). Values above~1
indicate late-stage diversification.

\paragraph{\texttt{centroid\_drift}.}
The Euclidean distance between the centroids of the early and late windows:
\begin{equation}
\label{eq:centroid}
  D_c
    = \bigl\lVert \bar{x}_{\mathrm{early}}
      - \bar{x}_{\mathrm{late}} \bigr\rVert.
\end{equation}
Large centroid drift~$D_c$ means the algorithm relocated its search
centre during the run it discovered a new promising region.  Small drift means the algorithm stayed
local throughout.

\paragraph{\texttt{f\_range\_early}, \texttt{f\_range\_late}, and
\texttt{f\_range\_ratio}.}
The range (maximum minus minimum) of raw fitness values in each window, and
their ratio:
\begin{align}
\label{eq:frange}
  F_{\mathrm{E}} &= \max_{t \le T/4} y_t
                    - \min_{t \le T/4} y_t, \\
  F_{\mathrm{L}} &= \max_{t > 3T/4} y_t
                    - \min_{t > 3T/4} y_t, \\
  R_F            &= F_{\mathrm{L}} \;/\; F_{\mathrm{E}}.
\end{align}
Here $F_{\mathrm{E}}$ and $F_{\mathrm{L}}$ are the early and late
fitness ranges, and $R_F$ is their ratio.
A shrinking fitness range ($R_F < 1$) indicates the algorithm is
converging
to a narrow fitness band near the optimum. A persistent or growing range
suggests the algorithm still samples points of widely varying quality.

% ============================================================================
\section{Novel Features}
\label{sec:features:novel}

The six features in this section are designed specifically for this thesis.  Each attempts to address a gap in the existing behavioural feature landscape that is not covered by prior defined metrics.  Together, they form one of the main contributions of this work.

\paragraph{\texttt{step\_size\_autocorrelation}.}
\label{sec:features:novel:ssac}

Autocorrelation of fitness values is a well-studied landscape descriptor
\cite{weinberger1990fitness}, and we include it above as
\texttt{fitness\_autocorrelation}.  But the
autocorrelation of \emph{step sizes}, measuring whether the algorithm's
movement magnitude is self-correlated over time, has not been applied
as a behavioural feature.

Let $s_t = \lVert x_{t+1} - x_t \rVert$ denote the step size at evaluation~$t$.
The lag-1 autocorrelation of the step-size series is:
\begin{equation}
\label{eq:ssac}
  \rho_s
    = \frac{\sum_{t=1}^{T-2}(s_t - \bar{s})(s_{t+1} - \bar{s})}
           {\sum_{t=1}^{T-1}(s_t - \bar{s})^2}.
\end{equation}

The step-size autocorrelation $\rho_s$ captures \emph{search momentum}. High positive autocorrelation (${\sim}0.9$) means the algorithm takes similarly-sized steps consecutively, characteristic of smooth step-size adaptation mechanisms such as the $1/5$-th success rule or CMA-ES's path-length control. Low or negative autocorrelation indicates erratic step-size variation, typical of random restarts or poorly
tuned mutation operators.

\paragraph{\texttt{fitness\_plateau\_fraction}.}
\label{sec:features:novel:plateau}

Existing stagnation metrics in BLADE (\texttt{longest\_no\_improvement\_streak},
\texttt{last\_improvement\_fraction}) measure plateaus in the \emph{best-so-far}
curve.  This conflates two distinct phenomena, the algorithm may be sampling diverse points that happen to be worse (active exploration), or it may be stuck in a genuinely flat region of the landscape.  We disentangle these by measuring plateaus in the \emph{raw} fitness sequence.

Let $\epsilon = 10^{-8} \cdot (\max_t y_t - \min_t y_t)$ be a relative
tolerance.  The plateau fraction is:
\begin{equation}
\label{eq:plateau}
  P_{\mathrm{flat}}
    = \frac{1}{T-1}\sum_{t=1}^{T-1}
      \mathbf{1}\!\bigl[|y_{t+1} - y_t| < \epsilon\bigr].
\end{equation}

High plateau fraction~$P_{\mathrm{flat}}$ means the algorithm repeatedly evaluates points with nearly identical fitness, indicating it is trapped in or deliberately exploiting a flat region.  This differs fundamentally from best-so-far stagnation, an algorithm can have a long no-improvement streak while still sampling diverse fitness values (because all are worse than the incumbent).  Conversely, a high plateau fraction with a short no-improvement streak indicates the algorithm has found a flat basin and is refining within it.

\paragraph{\texttt{half\_convergence\_time}.}
\label{sec:features:novel:halftime}

While \texttt{average\_convergence\_rate} captures the geometric mean of
per-step convergence ratios, it says nothing about when during the budget the
bulk of improvement occurs.  Two algorithms can have the same convergence rate
but very different temporal profiles. One may improve rapidly in the first 10\% of evaluations and then plateau, while another may start slowly but achieve a late breakthrough.

The half-convergence time measures when 50\% of the total best-so-far
improvement is reached:
\begin{equation}
\label{eq:halftime}
  t_{1/2}
    = \frac{1}{T-1}\min\!\bigl\{t : f^*_t \le f^*_1
      - \tfrac{1}{2}(f^*_1 - f^*_T)\bigr\}.
\end{equation}

Low values of the half-convergence time~$t_{1/2}$
(${\sim}0.001\text{--}0.01$) indicate rapid early convergence
followed by slow refinement. High values (${\sim}0.5\text{--}1.0$) indicate
a slow start with late breakthroughs.  The feature returns~$1.0$ when no
improvement occurs.

\paragraph{\texttt{improvement\_spatial\_correlation}.}
\label{sec:features:novel:impspatial}

No prior feature relates the \emph{distance} an algorithm travels to the
\emph{magnitude} of the improvement it achieves.  We define the improvement
spatial correlation as the Pearson correlation between step size and
improvement magnitude, restricted to improving steps:
\begin{align}
\label{eq:impspatial}
  \mathcal{I} &= \{t : y_t < f^*_{t-1}\}, \nonumber \\
  r_{\mathrm{IS}}
    &= r_{\mathrm{Pearson}}\!\bigl(
         \{s_{t-1}\}_{t \in \mathcal{I}},\;
         \{f^*_{t-1} - y_t\}_{t \in \mathcal{I}}
       \bigr).
\end{align}

Positive values of the improvement spatial
correlation~$r_{\mathrm{IS}}$ indicate that large jumps yield large
improvements.
Negative values indicate that small, local steps produce the biggest gains.  Values near zero mean improvement
magnitude is independent of the distance travelled.  This feature separates
algorithms that improve through directed long-range moves from those that
improve through local refinement.

\paragraph{\texttt{improvement\_burstiness}.}
\label{sec:features:novel:burst}

Barab\'asi~\cite{barabasi2005bursts} showed that many natural processes exhibit bursty dynamics. Events tend to cluster in time rather than arriving at a steady rate.
We apply this idea to the temporal pattern of fitness improvements.

Let $\mathcal{I} = \{t_1, t_2, \ldots, t_k\}$ be the sorted indices of
improving evaluations, and $\tau_i = t_{i+1} - t_i$ the inter-improvement
intervals.  The improvement burstiness is the coefficient of variation of these
intervals:
\begin{equation}
\label{eq:burst}
  B_{\mathrm{imp}}
    = \frac{\mathrm{std}(\tau)}{\mathrm{mean}(\tau)}.
\end{equation}

The improvement burstiness~$B_{\mathrm{imp}}$ is a coefficient of
variation (CV).  Values near~1 correspond to Poisson-like random
improvements.  CV $\gg 1$ indicates bursty dynamics where improvements arrive
in clusters separated by long stagnation phases.  CV $\ll 1$ indicates
metronomic, evenly-spaced improvements.  This feature characterises the
temporal \emph{texture} of the search process. Does the algorithm make steady
incremental progress, or does it alternate between productive bursts and idle
periods?

\paragraph{\texttt{dimension\_convergence\_heterogeneity}.}
\label{sec:features:novel:dimhet}

Standard convergence metrics (spread ratio, step-size trend) aggregate across
all dimensions.  They cannot detect whether the algorithm converges uniformly
in all dimensions or locks in on some dimensions while continuing to explore
others, a pattern that reflects sensitivity to problem separability.

For each dimension~$j$, we compute the range shrinkage ratio between the early
and late windows:
\begin{equation}
\label{eq:shrinkage}
  \sigma_j
    = \frac{\max_{t > 3T/4} x_t^{(j)} - \min_{t > 3T/4} x_t^{(j)}}
           {\max_{t \le T/4} x_t^{(j)} - \min_{t \le T/4} x_t^{(j)}}.
\end{equation}
The convergence heterogeneity is the standard deviation of these ratios across
dimensions:
\begin{equation}
\label{eq:dimhet}
  H_{\mathrm{dim}}
    = \mathrm{std}\!\bigl(\sigma_0, \sigma_1,
      \ldots, \sigma_{d-1}\bigr).
\end{equation}

Low dimension convergence heterogeneity~$H_{\mathrm{dim}}$ means all
dimensions converge at a similar rate, so the
algorithm treats the problem as isotropic.  High heterogeneity means some
dimensions converge tightly while others remain spread, suggesting the algorithm
identifies and exploits coordinate-wise structure.

% ============================================================================
\section{Computational Cost Analysis}
\label{sec:features:cost}

\Cref{tab:feature_cost} summarises the computational cost of each feature
category, measured over 50 repeated runs on a synthetic trajectory matching
the experimental setup ($T = 10{,}000$ evaluations, $d = 5$).  Each candidate
algorithm is evaluated on 50 MA-BBOB trajectories (10 problem instances
$\times$ 5 seeds), so per-candidate cost is $50\times$ the per-trajectory
cost.  The dominant cost is the existing BLADE metrics, which build a KD-tree
for dispersion and iterate over all evaluations for nearest-neighbour
distance.  The 17 features from categories 1, 4, and~5 add under 250\,ms per
candidate in total.

\begin{table}[t]
  \centering
  \caption{Computational cost of behavioural feature extraction.  Timings
    are the mean of 50 runs on a single CPU core for $T = 10{,}000$ evaluations
    in $d = 5$ dimensions.  SampEn uses subsampled trajectories ($N = 1{,}000$).}
  \label{tab:feature_cost}
  \footnotesize
  \begin{tabular}{@{}lrrl@{}}
    \toprule
    Category & \# & Per traj. (ms) & Per cand. \\
    \midrule
    Existing (BLADE)            & 11 & $743.7 \pm 5.3$   & $\sim$37\,s   \\
    Step-size dynamics          &  4 & $1.6 \pm 0.1$     & $\sim$80\,ms  \\
    Information-theoretic       &  4 & $27.5 \pm 0.5$    & $\sim$1.4\,s  \\
    Adapted pop.\ dynamics      &  7 & $1.0 \pm 0.1$     & $\sim$48\,ms  \\
    Novel features              &  6 & $1.5 \pm 0.1$     & $\sim$75\,ms  \\
    \midrule
    \textbf{Total}              & \textbf{32} & $\mathbf{775.2 \pm 5.4}$ & $\sim$\textbf{39\,s} \\
    \bottomrule
  \end{tabular}
\end{table}

Relative to the MA-BBOB evaluation itself (30--120\,s per candidate to run
the algorithm 50 times), the total feature extraction overhead of
${\sim}39$\,s per candidate is modest.  Over a full experimental run of 100
candidates, this amounts to roughly 65 minutes of feature extraction compared
to 1--3 hours of algorithm evaluation.
